Live coding

### 2018-12-06 wip-live-programming/paper.tex
Add section on live programming, working on calculus

\documentclass[sigplan,10pt]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}

\usepackage{lineno,hyperref,xcolor}
\usepackage{flushend}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{xypic}
\usepackage{semantic}
\usepackage{booktabs} 
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage[shortlabels]{enumerate}
\setlist{leftmargin=6mm}

\newcounter{thc}
\newcounter{dfc}

@@

\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
\acmYear{2017}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\setcopyright{none}
\bibliographystyle{ACM-Reference-Format}

\title{Design and implementation of a live coding environment for data science}

\author{Tomas Petricek}
%\affiliation{
%  \institution{The Alan Turing Institute}
%  \country{London, United Kingdom}
%}
%\email{tomas@tomasp.net}

\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{numclr}{rgb}{0.0,0.4,0.0}
\definecolor{strclr}{rgb}{0.4,0.4,0.0}
\definecolor{rstrclr}{rgb}{0.5,0.1,0.0}
\definecolor{prepclr}{rgb}{0.6,0.0,0.2}
\newcommand{\vect}[1]{\langl #1 \rangl}
\newcommand{\langl}{\begin{picture}(4.5,7)
\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\rangl}{\begin{picture}(4.5,7)
\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
\newcommand{\lsep}{~\,|\,~}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\rstr}[1]{\textnormal{\textcolor{rstrclr}{\sffamily "#1"}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\qident}[1]{\textnormal{\sffamily \textquotesingle #1\textquotesingle}}
\newcommand{\dom}{\ident{dom}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}

@@

\section{Understanding how data scientists work}

@@

Increasing numbers of data analysts use notebook systems such as Jupyter or R Markdown, which 
make it possible to combine text, equations and code with results of running the code such as tables 
or visualizations. Notebooks blur the traditional distinction between development and execution. 
Data analysts write small snippets of code, run them immediately to see results and then revise 
them interactively based on the results. We aim to take this trend further and build foundations
for seamlessly integrating development and execution in a notebook-like environment.

Jupyter notebooks are used by a variety of users ranging from scientists who implement complex
models to journalists who load data, perform simple aggregations and create visualizations. Our
focus is on simple use cases, such as those of journalists, because we believe that this is where
better tooling can provide the most value.

As an example, consider the data anlysis done by the Financial Times for an article on plastic 
waste. The analysis joins datasets from Eurostat, UN Comtrade and more, aggregates the data and 
builds a visualization comparing the plastic waste flows in 2017 and 2018.
Figure~\ref{fig:ft-uncomtrade} shows an extract from one notebook from the data analysis. 
The code of the analysis has a number of remarkable properties:
%
\begin{itemize}
\item There is no abstraction. The analysis uses lambda functions as arguments to library calls,
  but it does not define any functions. Parameterization is done either by defining a list of 
  inputs and then having a for loop, or by defining a top-level variable and setting its value.
  In our example, \ident{material} is set to \str{plastics} and a comment lists other options. 
  This lets the analyst easily see and check the results of intermediate steps of the analysis.

\item The code is structured as a sequence of commands. Some commands define a variable, either by
  loading data, or by transforming data loaded previously. Even in Python, data is often treated
  as immutable. Other commands produce an output that is displayed in the notebook.

\item There are many corner cases, such as the fact that the \ident{keep\_default\_na} parameter 
  needs to be set to handle Namibia correctly. These are discovered interactively by writing and 
  running a code snippet and examining the output, so providing a rapid feedback is essential.
\end{itemize}

Many Jupyter notebooks are more complex than the example discussed here. They might define
helper functions or even structure code using object-oriented programming. However, simple 
data analyses such as the one discussed here are frequent enough to pose an interesting and
important problem for programming tools research.

@@

\begin{figure}
\includegraphics[scale=0.52]{gamma.png}
\caption{The Gamma script to aggregate waste data. We select three countries and group
data by country. We then count the number of distinct reporting periods per country and type `.'
to choose further aggregation operations.}
\label{fig:gamma-waste}
\vspace{-1em}
\end{figure}

@@

@@

Data exploration of the kind discussed in the previous section has been the motivation
for a recent programming language research project The Gamma. The scripting language in The Gamma
captures the structure discussed above.

@@

Given the expressiveness of The Gamma, libraries for it are implemented in other languages, such as
JavaScript. More importantly, The Gamma supports type providers, which can be used to generate types,
on the fly, for accessing and exploring external data sources. Type providers provide object types 
with members and The Gamma makes using those convenient by providing auto-complete when the user 
types the dot symbol (`.') to access a member.

The combination of type providers and auto-complete on~``.'' makes it possible to complete a 
large number of data exploration tasks through a very simple interaction of selecting operations from
a list. An example in Figure~\ref{fig:gamma-waste} summarizes waste data using a type provider.
The names of identifiers such as \qident{average Waste (kg)} do not have
any special meaning -- they are merely members generated by the type provider, based on information
about the data source. The type provider used in this example generates an object with members for
different data transformations, such as \qident{group data}, which return furhter objects with 
members for specifying transformation parameters, such as selecting the grouping key using
\qident{by Country}.

@@

\section{Calculus}

x

\begin{equation*}
\begin{array}{lcl}
p &=& c_1; \ldots; c_n\\
c &=& \kvd{let}~x = e ~|~ e\\
e &=& n ~|~ x ~|~ e.m(a, \ldots, a)\\
a &=& e ~|~ \lambda x\rightarrow e
\end{array}
\end{equation*}

>>>>

\documentclass[acmsmall,anonymous,fleqn]{acmart}\settopmatter{printfolios=false,printccs=false,printacmref=false}

%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{OOPSLA}
\acmArticle{1}
\acmYear{2019}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

\usepackage{booktabs}
\usepackage{subcaption}
\usepackage{lineno,hyperref,xcolor}
\usepackage{flushend}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{xypic}
\usepackage{semantic}
\usepackage{booktabs} 
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage{enumerate}
\usepackage{amsmath}

\setlist{leftmargin=6mm}
\newcounter{thc}
\newcounter{dfc}

@@

\title{Live coding}
\author{Tomas Petricek}

\affiliation{
  \institution{University of Kent}
  \country{United Kingdom}
}
\email{tomas@tomasp.net}

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}

\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{numclr}{rgb}{0.0,0.4,0.0}
\definecolor{strclr}{rgb}{0.4,0.4,0.0}
\definecolor{rstrclr}{rgb}{0.5,0.1,0.0}
\definecolor{prepclr}{rgb}{0.6,0.0,0.2}
\newcommand{\vect}[1]{\langl #1 \rangl}
\newcommand{\langl}{\begin{picture}(4.5,7)
\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\rangl}{\begin{picture}(4.5,7)
\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
\newcommand{\lsep}{\;\;|\;\;}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\rstr}[1]{\textnormal{\textcolor{rstrclr}{\sffamily "#1"}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\qident}[1]{\textnormal{\sffamily \textquotesingle #1\textquotesingle}}
\newcommand{\dom}{\ident{dom}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}

@@

\section{Understanding how data scientists work}
\label{sec:background}

@@

Data analysts increasingly use notebook systems such as Jupyter or R Markdown, which 
make it possible to combine text, equations and code with results of running the code such as tables 
or visualizations. Notebooks blur the distinction between development and execution. 
Data analysts write small snippets of code, run them to see results immediately and then revise 
them interactively based on the results. We aim to build foundations for such seamless integration 
of development and execution in a notebook environment.

Jupyter notebooks are used by a variety of users ranging from scientists who implement complex
models to journalists who load data, perform simple aggregations and create visualizations. Our
initial focus is on simple use cases, such as those of journalists, because we believe that this 
is where better tooling can provide the most value.

As an example, consider the data anlysis done by the Financial Times for an article on plastic 
waste. The analysis joins datasets from Eurostat, UN Comtrade and more, aggregates the data and 
builds a visualization comparing the plastic waste flows in 2017 and 2018.
Figure~\ref{fig:ft-uncomtrade} shows an extract from one notebook from the data analysis. 
The code of the analysis has a number of remarkable properties:
%
\begin{itemize}
\item There is no abstraction. The full analysis uses lambda functions as arguments to library calls,
  but it does not define any top-level functions. Parameterization is done by defining a list of 
  inputs and then having a for loop, or by defining a top-level variable and setting its value.
  In our example, \ident{material} is set to \str{plastics} and a comment lists other options. 
  This lets the analyst easily see and check the results of intermediate steps of the analysis.

\vspace{0.5em}
\item The code is structured as a sequence of commands. Some commands define a variable, either by
  loading data, or by transforming data loaded previously. Even in Python, data is often treated
  as immutable. Other commands produce an output that is displayed in the notebook.

\vspace{0.5em}
\item There are many corner cases, such as the fact that the \ident{keep\_default\_na} parameter 
  needs to be set to handle Namibia correctly. These are discovered interactively by writing and 
  running a code snippet and examining the output, so providing a rapid feedback is essential.
\end{itemize}

Many Jupyter notebooks are more complex than the example discussed here. They might define
helper functions or even structure code using object-oriented programming. However, simple 
data analyses such as the one discussed here are frequent enough to pose an interesting and
important domain for programming tools research.

@@

@@

\begin{figure}
\includegraphics[scale=0.28]{gamma1.png}
\caption{\small The Gamma script to aggregate Olympic medals. We select Rio 2016 games and
count the number of distinct events per athlete. We then type `.' to choose 
further aggregation operations.}
\label{fig:gamma-dot}
\vspace{-1em}
\end{figure}

@@

Data exploration of the kind discussed in the previous section has been the motivation
for a recent programming environment called The Gamma. The scripting language in The Gamma
captures the structure discussed above.

@@

Given the limited expressiveness of The Gamma, libraries for it are implemented in other languages, such as
JavaScript. More importantly, The Gamma supports type providers, which can be used to generate types,
on the fly, for accessing and exploring external data sources. Type providers provide object types 
with members and The Gamma makes using those convenient by providing auto-complete when the user 
types the dot symbol (`.') to access a member.

The combination of type providers and auto-complete on~``.'' makes it possible to complete a 
large number of data exploration tasks through a very simple interaction of selecting operations from
a list. An example in Figure~\ref{fig:gamma-dot} summarizes data on Olympic medals using a type provider.
The names of identifiers such as \qident{sum Bronze} do not have
any special meaning -- they are merely members generated by the type provider, based on information
about the data source. The type provider used in this example generates an object with members for
different data transformations, such as \qident{group data}, which return furhter objects with 
members for specifying transformation parameters, such as selecting the grouping key using
\qident{by Athlete}.

@@

% https://www.ft.com/content/360e2524-d71a-11e8-a854-33d6f82e62f8
% https://github.com/ft-interactive/recycling-is-broken-notebooks

\begin{figure}
\includegraphics[scale=0.28 ]{gamma2.png}
\caption{\small Live preview in The Gamma programming environment. The table is produced as
  the data anlyst types and updates on-the-fly to show result at the current cursor position.}
\label{fig:gamma-preview}
\end{figure}

\section{Live coding for data exploration}

This paper presents the foundations of a live coding environment for data exploration that 
follows the analysis in the previous section. We consider simple data exploration tasks, such as 
those done by journalists and we implement our ideas in the editor for TheGamma.

Figure~\ref{fig:gamma-preview} shows a TheGamma script exporing Olympic medal statistics in 
our live coding environment. Before discussing the technical details, it is worth highlighting
a number of important points about our approach.

\paragraph{Real-world use cases.}
Our work has been motivated by the needs outlined in the previous section. In particular, we 
have been focusing on a langauge that is simple, yet sufficient for completing basic data 
exploration tasks that a journalist might want to do. We have used our live coding environment
to perform a number of simple data analyses looking at topics such as public government spending, 
crises in financial markets and activities of a research institute. Qualitative observations about
our experience are reported in Section~X. Our tools can be used in any modern web browser and are 
available as an open-source package.

\paragraph{Text editor integration.}
Our approach is based on an ordinary text editor that allows unrestricted modification of the
code. Our parser supports error-recovery and we attempt to give previews whenever the code can be
correctly parsed and type-checked up to the current location. Notably, we do not, in general, know
how the user modified the code. This is in contrast with approaches based on structured editing
where the editor tracks what changes to code have been made. Using an ordinary text editor makes
providing live previews harder, but we believe that many users still favor the flexibility of 
plain text editors.

\paragraph{Data stay, code changes.}
Our focus has been on the scenario when input data is a static file that does not change, but the
data analyst modifies the source code. This is in contrast with work on incremental computation,
which focuses on scenarios where code stays the same, but input data change.

That said, many of the methods that we used for evaluating live previews such as maintaining 
the dependency graph with nodes for individual syntactic elements are closely related to methods
used in incremental computation research. We do not attempt to combine the two, but note that
this is an interesting future direction.

\newpage
\begin{figure}
\raggedright
{\small\sffamily Programs, commands, expressions, arguments and values}
\begin{equation*}
\hspace{-2em}
\begin{array}{lcl}
p &=& c_1; \ldots; c_n\\
c &=& \kvd{let}~x = e \lsep e\\
\end{array}
\quad
\begin{array}{lcl}
e &=& o \lsep x \\
  &|& e.m(a, \ldots, a)
\end{array}
\quad
\begin{array}{lcl}
a &=& e \lsep \lambda x\rightarrow e\\
v &=& o \lsep \lambda x\rightarrow e\\
\end{array}
\end{equation*}

\vspace{0.5em}
{\small\sffamily Evaluation contexts of expressions}
\begin{equation*}
\hspace{-2em}  
\begin{array}{lcl}
C_e[-] &=& -.m(a_1, \ldots, a_n) ~|~ -.m(v_1, \ldots, v_{m}, -, a_1,\ldots , a_n) ~|~ -\\
C_c[-] &=& \kvd{let}~x = C_e[-] ~|~ C_e[-]\\
C_p[-] &=& c_1;\; \ldots;\; c_{k-1};\; C_c[-];\; c_{k+1};\; \ldots;\; c_n\\
\end{array}
\end{equation*}

\vspace{0.5em}
{\small\sffamily Let elimination and member reduction}
\begin{equation*}
\hspace{-2em}  
\begin{array}{ll}
\hspace{-0.5em}\begin{array}{l}  
c_1;\; \ldots;\; c_{k-1};\; \kvd{let}~x=o;\; c_{k+1};\; \ldots; c_n \rightsquigarrow\\
\qquad  c_1;\; \ldots;\; c_{k-1};\; c_{k+1}[x\leftarrow o];\; \ldots; c_n[x\leftarrow o]\end{array} &
\quad\text{\small\sffamily(let)}\\[1.2em]
{o.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o'}\implies
  {C_p[o.m(v_1, \ldots, v_n)] \rightsquigarrow C_p[o']} & 
\quad\text{\small\sffamily (external)}    
\end{array}  
\end{equation*}

\vspace{-0.5em}
\caption{Syntax, contexts and reduction rules of the data exploration calculus}
\label{fig:dec-calculus}
\end{figure}

\section{Data exploration calculus}
\label{sec:calculus}

The \emph{data exploration calculus} is a small, formally tractable language for data exploration. 
The calculus is not Turing-complete and it can only be used together with external libraries that
define what objects are available and what the behaviour of their members is. However, this is 
sufficient to capture the simple data analyses discussed in Section~\ref{sec:background}. We 
define the caluclus in this section and then use it as the basis for discussing our live preview 
mechanism in Section~\ref{sec:formal}.

\subsection{Language syntax}
The calculus combines object-oriented features such as member access with functional features
such as lambda functions. The syntax is defined in Figure~\ref{fig:dec-calculus}. We write $x$ for 
variables and $o$ for object values. A program $p$ in the data exploration calculus consists of a 
sequence of commands $c$. A command can be either let binding or an expression. Lambda functions 
can only appear directly as arguments in method calls. An argument $a$ can thus be a lambda 
function or an expression.

\subsection{Operational semantics}
The data exploration calculus is a call-by-value language. We model the evaluation as a small-step
reduction $\rightsquigarrow$. Fully evaluating a program results in an irreducible sequence of 
objects $o_1;\; \ldots;\; o_n$ which can be displayed as results of the data analysis. The 
operational semantics is parameterized by a relation $\rightsquigarrow_\epsilon$ that models
the functionality of the external libraries used with the calculus and defines reduction rules
for member accesses. The relation has the following form:
%
\begin{equation*}
o.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o'
\end{equation*}
%
foo

bar

\subsection{Equational theory}
In The Gamma language, computations are expressed using
member access, written as $e.m(e_1,\ldots, e_n)$. In this paper, we do not define how member
access evaluates. This has been done elsewhere \cite{gamma}, but more importantly, the evaluation of
previews does not rely on the exact specifics of the evaluation, provided that the language
satisfies certain basic conditions. The following definitions provides the necessary structure 
for discussing correctness of previews.
